---
title: "02_las_to_dem"
author: "Matthew Coghill"
date: "11/30/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The primary purpose of this file will be to process the downloaded LAS files into raster products.

```{r}
# ☻

library(tidyverse)
library(lidR)
library(future)
library(terra)

```

Get directories and load the las file as a catalogue

```{r}

dl_dir <- file.path("./01_download_raw")
las_file <- readLAScatalog(dl_dir)
tile_dir <- file.path("./02_tile")
prod_dir <- file.path("./outputs")
dir.create(tile_dir, showWarnings = FALSE)
dir.create(prod_dir, showWarnings = FALSE)

# Perform tiling
plan(multisession)
set_lidr_threads(1)

# Catalog processing options
opt_chunk_buffer(las_file) <- 0
opt_chunk_size(las_file) <- 250
opt_output_files(las_file) <- file.path(tile_dir, "{XLEFT}_{YBOTTOM}")
opt_chunk_alignment(las_file) <- c(
  plyr::round_any(xmin(las_file), 10, floor),
  plyr::round_any(ymin(las_file), 10, floor))

ctg <- catalog_retile(las_file)
ctg <- readLAScatalog(tile_dir)
lidR:::catalog_laxindex(ctg)

```

Some point classification has been done but I am not thoroughly satisfied with ground point classification. Reclassify ground points here:

```{r}

# Create function
ctg_fn <- function(las) {
  las <- readLAS(las)
  if (is.empty(las)) return(NULL)  
  l1 <- filter_poi(las, !Classification %in% c(0L, 2L))
  l2 <- filter_poi(las, Classification %in% c(0L, 2L))
  l2$Classification <- 0L
  l2 <- classify_ground(l2, algorithm = csf())
  las <- rbind(l1, l2)
  las <- filter_poi(las, buffer == 0)
  return(las)
}

class_out <- file.path("./02a_classified")
dir.create(class_out, showWarnings = FALSE)

opt_chunk_buffer(ctg) <- 10
opt_output_files(ctg) <- file.path(class_out, "{*}")

# Function is a memory hog, use only half the amount of cores to limit number
# of tiles loaded in at once
plan(multisession(workers = availableCores()/2))
set_lidr_threads(1)
ctg_cls <- catalog_apply(ctg, ctg_fn, .options = list(automerge = TRUE))
plan(multisession)
lidR:::catalog_laxindex(ctg_cls)

```

Digital terrain/elevation model (DTM/DEM) generation: use tin algorithm. Luckily, the points are already classified and noise has been removed, so we can just run the algorithm!

```{r}

rast_out <- file.path("./03_dem_tiles")
dir.create(rast_out, showWarnings = FALSE)

# Set catalog options
opt_chunk_buffer(ctg_cls) <- 10
opt_output_files(ctg_cls) <- file.path(rast_out, "{*}_dem")

# Create function so that the output is not a large raster file
set_lidr_threads(1)
plan(multisession(workers = availableCores()/2))
dem_5cm <- grid_terrain(ctg_cls, res = 0.05, algorithm = tin())
plan(sequential)
writeRaster(dem_5cm, file.path(prod_dir, "dem_5cm.tif"), overwrite = TRUE)

```

Now, produce DSM (don't need to normalize here)

```{r}

dsm_dir <- file.path("./05_dsm_tiles")
dir.create(dsm_dir, showWarnings = FALSE)

ctg_cls <- readLAScatalog(class_out)
opt_chunk_buffer(ctg_cls) <- 10
opt_output_files(ctg_cls) <- file.path(dsm_dir, "{*}_dsm")

plan(multisession(workers = availableCores()/2))
set_lidr_threads(1)

dsm <- grid_canopy(ctg_cls, 0.05, dsmtin())
plan(sequential)
writeRaster(dsm, file.path(prod_dir, "dsm_5cm.tif"), overwrite = TRUE)

```

Post-process the DSM and then Clip the DSM area to match the DEM

```{r}

dem <- rast(file.path(prod_dir, "dem_5cm.tif"))
dsm <- rast(file.path(prod_dir, "dsm_5cm.tif"))

fill.na <- function(x, i=5) { if (is.na(x)[i]) { return(mean(x, na.rm = TRUE)) } else { return(x[i]) }}
w <- matrix(1, 3, 3)

filled <- focal(dsm, w, fun = fill.na)
# smoothed <- focal(chm, w, fun = mean, na.rm = TRUE)

dsm_mask <- mask(filled, dem, filename = file.path(prod_dir, "dsm_5cm_clip.tif"), overwrite = TRUE)

```


For CHM, we need normalized LAS files - CHM not required so far...

```{r}

# norm_dir <- file.path("./04_normalized")
# dir.create(norm_dir, showWarnings = FALSE)
# 
# ctg_cls <- readLAScatalog(class_out)
# opt_chunk_buffer(ctg_cls) <- 10
# opt_output_files(ctg_cls) <- file.path(norm_dir, "{*}")
# 
# plan(multisession(workers = 6))
# set_lidr_threads(2)
# ctg_norm <- normalize_height(ctg_cls, algorithm = tin(), Wdegenerated = FALSE)
# lidR:::catalog_laxindex(ctg_norm)

```

Now, create CHM

```{r}

# ctg_norm <- readLAScatalog(norm_dir, filter = "-drop_z_below 0")
# max_z <- quantile(ctg_norm$Max.Z[-which.max(ctg_norm$Max.Z)], 0.9)
# sq <- c(0, 2, seq(5, plyr::round_any(max_z, 5, ceiling), 5))
# 
# rast_out <- file.path("./05_dsm_tiles")
# dir.create(rast_out, showWarnings = FALSE)
# 
# # Set catalog options
# opt_chunk_buffer(ctg_norm) <- 10
# opt_output_files(ctg_norm) <- file.path(rast_out, "{*}_dsm")
# 
# # Create function so that the output is not a large raster file
# plan(multisession(workers = 6))
# set_lidr_threads(2)
# chm_5cm <- grid_canopy(ctg_norm, res = 0.05, algorithm = pitfree(sq, c(0, 0.5), subcircle = 0.025))
# plan(sequential)
# writeRaster(chm_5cm, file.path(prod_dir, "chm_5cm.tif"), overwrite = TRUE)

```
